\section{Introduction}

Time series missingness is usually not homogeneous. In practical data collection systems, random sensor dropout, mechanism-driven suppression, and structured co-missing events can appear simultaneously. This leads to mixed settings where MAR-like and MNAR-like missing positions coexist within one sequence. Under such conditions, applying a single uniform imputation rule is often insufficient.

Recent imputation baselines have significantly improved recovery quality, including recurrent models, attention-based models, graph-temporal models, and generative denoising models \cite{cao2018brits,du2023saits,liu2024imputeformer,marisca2022spin,wu2023timesnet,chen2024fgti,liu2024mtsci}. However, based on our local benchmark review and CIKM draft positioning, most prior methods still couple multiple dependency types in a shared backbone: observed-to-observed context (OO), observed-to-missing interaction (OM), and missing-to-missing structure (MM). In particular, MM is rarely promoted to an explicit branch with dedicated routing and interpretation.

This work targets two concrete gaps. First, existing pipelines often treat missingness as random noise and under-model missing-side structure. Second, in mixed MAR/MNAR scenarios, different missing positions may require different recovery dependencies; static fusion cannot reliably reflect this variation.

To address these issues, we propose UniMiss with an explicit decomposition:
\begin{itemize}
\item \textbf{OO foundational context path}: extracts stable observed-side context and decouples it from missing-side decision making.
\item \textbf{OM interaction-oriented expert branch}: performs prompt-conditioned routing over observed evidence experts for each target missing position.
\item \textbf{MM missing-structure expert branch}: explicitly models co-missing topology, periodic evidence, and extreme-event clues via missing-side experts.
\item \textbf{Stage-II mechanism-aware soft gate}: uses mechanism prompt and local/global missing summaries to adaptively balance OM and MM under mixed missingness.
\end{itemize}

The paper draft is tightly aligned with the repository implementation scope: (1) baseline-consistent data protocol, (2) modular ablation switches for OO/OM/MM/Stage-II, (3) auxiliary studies including hyperparameter, visualization, and scalability tracks, and (4) ACM CIKM-style manuscript structure.

Our main contributions are summarized as follows:
\begin{itemize}
\item We provide a mechanism-aware architecture that explicitly separates OO, OM, and MM responsibilities instead of relying on a monolithic imputation backbone.
\item We design a Stage-II MAR-vs-MNAR soft gate for position-wise adaptive fusion between OM and MM in mixed missingness settings.
\item We build a unified experimental protocol and code interface with switchable module-level ablations and reproducible analysis hooks.
\end{itemize}
